

## Direct weighing for certain compartments and volume and density measurements for others

The second case we consider here is that where felling constraints mean that only semi-destructive measurements are taken. These combine direct measurements for certain parts of the tree, and volume and density measurements for the others. We will illustrate this by developing an allometric equation for the dry northern forests of Cameroon. The biomass of dry forests is particularly difficult to determine given the architectural complexity of the trees they contain. Human intervention is also particularly significant in dry areas because of the rarity of forest resources and the magnitude of the demand for bioenergy. This is reflected by the trimming, pruning and maintenance practices often employed for trees located in deciduous forests, agro-forest parks and hedges (photo \@ref(fig:ka)).

```{r ka, out.width='100%', fig.scap="Trimming shea trees (*Vitellaria paradoxa*) in north Cameroon", fig.cap="(ref:ka)"}

ka

```

The trees in most dry areas are protected as this woody resource regenerates particularly slowly and is endangered by human activities. Biomass measurements are therefore non-destructive and take full advantage of this trimming to measure the biomass of the trimmed compartments. Grazing activities limit regeneration and small trees are often few in number. This part of the guide will therefore consider only mature trees.


### In the field: case of semi-destructive measurements

Generally, the trunk and the large branches are not trimmed; only the small branches are affected. The measurement of fresh biomass (in kg) may be divided into two parts: measuring trimmed fresh biomass and measuring untrimmed fresh biomass (Figure \@ref(fig:emond)A).


```{r emond, fig.align='center', fig.cap="(ref:emond)", fig.process=pdf2png, fig.scap="Determination of total fresh biomass", cache=TRUE, dev='tikz', out.width='100%'}

#invisible(plot.new())
grid.arrange(
  white_space, 
  pictureGrob(emonde1), 
  rasterGrob(emonde2, x = 0.5, y = 0.5, w = 0.7), 
  ncol = 3, widths = unit(c(8,7,6), "cm")#, heights = unit(8, "cm") 
  )
grid.text("Fresh biomass of small untrimmed",
          x = 0.01, y = 0.8, just = 'left')
grid.text("branches calculated from their basal",
          x = 0.01, y = 0.76, just = 'left')
grid.text("circumference and a biomass table",
          x = 0.01, y = 0.72, just = 'left')

grid.text("Fresh biomass of large untrimmed",
          x = 0.01, y = 0.56, just = 'left')
grid.text("branches and trunk, calculated from",
          x = 0.01, y = 0.52, just = 'left')
grid.text("volume and density measurements.",
          x = 0.01, y = 0.48, just = 'left')
grid.text("Hypotheses: the sections cut are",
          x = 0.01, y = 0.44, just = 'left')
grid.text("considered to be cylinders and density",
          x = 0.01, y = 0.4, just = 'left')
grid.text("is considered to be the same in all the",
          x = 0.01, y = 0.36, just = 'left')
grid.text("compartments of the tree",
          x = 0.01, y = 0.32, just = 'left')

grid.text("Fresh biomass of the trimmed",
          x = 0.01, y = 0.16, just = 'left')
grid.text("branches, measured by weighing",
          x = 0.01, y = 0.12, just = 'left')

grid.text("(A)",
          x = 0.4, y = 0.85, just = 'left')
grid.text("(B)",
          x = 0.8, y = 0.85, just = 'left')

```


#### Trimmed fresh biomass {-}

The branches may be trimmed in compliance with local practices (often using a machete). Use a measuring tape to determine the diameter at the base of each branch. Then separate the leaves from the trimmed branches. Determine (by weighing separately) the fresh biomass of the leaves from the trimmed branches ($B_{\text{trimmed fresh leaf}}$) and the fresh biomass of the wood from the trimmed branches ($B_{\text{trimmed fresh wood}}$). Suitable scales should be used for these weighing operations. If the leaves weigh less than 2 kg, they may be weighed using field-work electronic scales.

Take a random sample of the leaves from the trimmed branches. At least three samples of leaves from three different branches are generally required to constitute the aliquot. Measure its fresh weight ($B_{\text{fresh leaf}}^{\text{aliquot}}$ in g). Take also an aliquot of the wood at random from the trimmed branches, without debarking. Measure its fresh mass ($B_{\text{fresh wood}}^{\text{aliquot}}$ in g) in the field, immediately after cutting. Place these aliquots in numbered plastic bags and send to the laboratory. The fresh volume of the wood aliquot will be measured later in the lab (see \S\ \@ref(lab2)), and the value used to determine mean wood density $\bar{\rho}$.


#### Untrimmed fresh biomass {-}

Untrimmed biomass is measured indirectly as non-destructive. The different branches in the trimmed tree must first be numbered (Figure \@ref(fig:emond)B). The small untrimmed branches must be processed differently from the large branches and the trunk (Figure \@ref(fig:emond)A). For the small branches, only basal diameter needs to be measured. The biomass of these small untrimmed branches is estimated from the relationship between their basal diameter and their mass, as explained in section \@ref(calBsp).

The biomass of the trunk and the large branches is estimated from measurements of volumes ($V_i$ in cm^3^) and mean wood density ($\bar{\rho}$ in g\ cm^-3^). The large branches and trunk should be divided virtually into sections that are then materialized by marking the tree. The volume $V_i$ of each section $i$ is obtained by measuring its diameter (or its circumference) and its length. Sections about one meter in length are preferable in order to consider diameter variations along the length of the trunk and branches.


### In the laboratory {#lab2}

First, measure the volume ($V_{\text{fresh wood}}^{\text{aliquot}}$) of the wood aliquot taken from the trimmed compartments. This volume may be measured in different manners [@maniatis11]. That most commonly employed measures the volume of water displaced when the sample is immersed in water. The volume of water displaced may be measured using a graduated tube of suitable dimensions for the sample (Figure \@ref(fig:archim)). Another method consists in cutting the sample into shapes whose volume may be accurately measured. This method requires precision tools and personnel trained in cutting wood.

```{r archim, out.width='100%', fig.cap="(ref:archim)"}

knitr::include_graphics(paste0(fig_path, "/archim.pdf"))

```

The wood and leaf aliquots should then be subject to the same laboratory measurements (oven drying, determination of dry weight, etc. ) as described in section \@ref(lab1).


### Calculations

The dry biomass of the tree is obtained by the sum of the trimmed dry biomass and the untrimmed dry biomass:

\[
B_{\text{dry}}=B_{\text{trimmed dry}}+B_{\text{untrimmed dry}}
\]


#### Calculating trimmed biomass {-}

From the fresh biomass $B_{\text{fresh wood}}^{\text{aliquot}}$ of a wood aliquot and its dry biomass $B_{\text{dry wood}}^{\text{aliquot}}$, calculate as above (see equation \@ref(eq:qbois)) the moisture content of the wood (including bark):

\[
\chi_{\text{wood}}=\frac{B^{\text{aliquot}}_{\text{dry wood}}}{B^{\text{aliquot}}_{\text{fres wood}}}
\]

Likewise, calculate the moisture content of the leaves from the fresh biomass $B_{\text{fresh leaf}}^{\text{aliquot}}$ of the leaf aliquot and its dry biomass $B_{\text{dry leaf}}^{\text{aliquot}}$:

\[
\chi_{\text{leaf}}=\frac{B^{\text{aliquot}}_{\text{dry leaf}}}{B^{\text{aliquot}}_{\text{fresh leaf}}}
\]

Trimmed dry biomass can then be calculated:

\[
B_{\text{trimmed dry}}=B_{\text{trimmed fresh wood}}\times \chi_{\text{wood}}+B_{\text{trimmed fresh leaf}}\times \chi_{\text{leaf}}
\]

where $B_{\text{trimmed fresh leaf}}$ is the fresh biomass of the leaves stripped from the trimmed branches and $B_{\text{trimmed fresh wood}}$ is the fresh biomass of the wood in the trimmed branches.


#### Calculating untrimmed biomass {#calBsp}

Two calculations are required to calculate the dry biomass of the untrimmed part (i.e. that still standing): one for the small branches, the other for the large branches and the trunk. The untrimmed biomass is the sum of the two results:

\[
B_{\text{untrimmed dry}}=B_{\text{untrimmed dry branch}}
+B_{\text{dry section}}
\]

Each section $i$ of the trunk and the large branches may be considered to be a cylinder of volume (Smalian's formula):

\begin{equation}
V_i=\frac{\pi}{8}\ L_i\ (D_{1i}^2+D_{2i}^2)(\#eq:cyl)
\end{equation}

where $V_i$ is the volume of the section $i$, $L_i$ its lenght, and $D_{1i}$ and $D_{2i}$ are the diameters of the two extremities of section $i$. The truncated cone volume formula (see equation \@ref(eq:cone)) can also be used instead of the cylinder formula (\@ref(eq:cyl)), but the difference between the results will be slight as the tapering over one meter is not very pronounced in trees.

The dry biomass of the large branches and trunk is the product of mean wood density and total volume of the large branches and trunk:

\begin{equation}
B_{\text{dry section}}=\bar{\rho}\times\sum_iV_i(\#eq:Btr)
\end{equation}

where the sum corresponds to all the sections in the large branches and the trunk (Figure \@ref(fig:emond)B), and where mean wood density is calculated by:

\[
\bar{\rho}=\frac{B_{\text{dry wood}}^{\text{aliquot}}}
{V_{\text{fresh wood}}^{\text{aliquot}}}
\]

Care must be taken to use consistent measurement units. For example, if mean wood density $\bar{\rho}$ in (\@ref(eq:Btr)) is expressed in g\ cm^-3^, then volume $V_i$ must be expressed in cm^3^, meaning that both length $L_i$ and diameters $D_{1i}$ and $D_{2i}$ must also be expressed in cm. Biomass in this case is therefore expressed in g.

The dry biomass of the untrimmed small branches should then be calculated using a model between dry biomass and basal diameter. This model is established by following the same procedure as for the development of an allometric model (see chapters \@ref(don) to \@ref(util) herein). Power type equations are often used:

\[
B_{\text{dry branch}}=a+bD^c
\]

where $a$, $b$ and $c$ are model parameters and $D$ branch basal diameter, but other regressions may be tested (see \@ref(tab:prim)). Using a model of this type, the dry biomass of the untrimmed branches is:

\[
B_{\text{untrimmed dry branch}}=\sum_j(a+bD_j^c)
\]

where the sum is all the untrimmed small branches and $D_j$ is the basal diameter of the branch $j$.

<!--------------------->
<!-- END SECTION 3.2 -->
<!--------------------->
