
## Measuring roots

It is far more difficult to determine root biomass than above-ground biomass. The methods we propose here are the results of campaigns conducted in different ecosystems and were compared in a study conducted in Congo [@levillain11].

The first step, whatever the ecosystem, consists in drawing a Voronoi^[A Voronoi diagram (also called a Voronoi decomposition, a Voronoi partition or Voronoi polygons) is a special kind of decomposition of a metric space determined by distances to a specified family of objects, generally a set of points.] 
diagram around the selected tree. Figure \@ref(fig:voro) indicates the procedure employed: *(i)* draw the segments that link the selected tree to each of its neighbors; *(ii)* draw the bisectors for each segment, *(iii)* connect the bisectors one to another to bound a space around the tree; *(iv)* this space can then be divided into joined triangles, the area of each zone being easy to calculate when the length of all three sides ($a$, $b$ and $c$) is known:

\[
A=\sqrt{p(p-2a)(p-2b)(p-2c)}
\]

where $p=a+b+c$ is the triangle perimeter and $A$ its area. Figure \@ref(fig:vanu) illustrates this procedure for coconut plantations in Vanuatu [@navarro08].

```{r voro, fig.cap="(ref:voro)"}

knitr::include_graphics(paste0(fig_path, "/voronoi.pdf"))

```

```{r vanu, fig.scap="Typical Voronoi diagram used for root sampling in a coconut plantation in Vanuatu", fig.cap="(ref:vanu)"}

#plot.new()
grid.picture(vanu)
grid.raster(photo_list$cocoteraie, x = 0.1, y = 0.1, w = 0.5)

```

The space established in this manner is not a materialization of the tree's "vital" space. It is simply a manner used to cut space into joined area to facilitate the sampling of below-ground biomass. The major hypothesis here is that the roots of other trees entering this space compensate for those of the selected tree leaving this space.

In the case of multispecific stands or agro-forest, it is sometimes impossible to separate the roots of the different species. In this case its would be very risky to construct individual models (root biomass related to the sampled tree), but root biomass estimations per hectare, without any distinction between species, are perfectly valid.

Sampling methods vary with root size. @levillain11 conducted a study to compare different methods in the same tree (photo \@ref(fig:evor)). They showed that the most profitable, in terms of cost-precision, was to sample the finest roots by coring, whereas medium-size roots require partial excavation and large roots total excavation from the Voronoi space.

```{r evor, fig.scap = "Superimposition of different sampling methods (cores, excavations by cubes, partial Voronoi excavation, total Voronoi excavation)", fig.cap="(ref:evor)"}

evor

```
\begin{photo}
\includetwophotos{root1}{0.5010309}{root2}{0.4989691}%% root1: 2592 x 1936  root2: 3072 x 2304
\caption[]{label{evor}}
\end{photo}

The number of cores obtained and the size of the pit dug vary from one ecosystem to another. In Congo, in eucalyptus plantations, the optimum number of cores for 10\ \% precision is about 300 on the surface (0--10~cm) and about 100 for deeper roots (10--50 and 50--100~cm). Such a level of precision to a depth of 1~m requires 36 man-days. By contrast, if only 30\ \% precision is sought, this cuts sampling time by 75\ \%. This example clearly illustrates the usefulness of pre-sampling (see chapitre \ref{samp}) in order to evaluate the variability of the studied ecosystem then adapt the protocol to goals and desired precision.

Once the soil samples with the roots have been taken, the cores containing the fine roots can be sorted in the laboratory. On the other hand, given the volume and weight of the earth excavated, the medium-size and large roots must be sorted in the field. In the laboratory, the soil sample should first be washed through a filter and roots recovered either by flotation and/or by sieving. In the field, the samples should be sorted manually on tarpaulins. An air knife can be used on the large and medium-size roots to excavate the root system fully while preserving its architecture. This method, which is particularly advantageous in sandy soils, helps meet biomass and architecture goals but does require the presence of a mobile compressor (photo \ref{air}).

\begin{photo}
\setlength{\unitlength}{\textwidth}%
\begin{picture}(1,0.2926829268292683)
\put(0,0){\includethreephotos{air1}{0.3902439024390244}{air2}{0.2195121951219512}{air3}{0.39024390243902449}}
\put(0.5,0){\includegraphics[width=0.108\textwidth]{air4}}
\end{picture}
\caption[Air knife being used in Congo to extract the root systems of eucalyptus.]{Air knife being used in Congo to extract the root systems (large and medium-size roots) of eucalyptus. Left, operator with protective equipment (dust, noise); middle, compressor and close up of the air pressure gauge (about 8 bars); right, the result (photos: C. Jourdan).\label{air}}
\end{photo}

Once the roots have been sorted and harvested they should be dried in an oven using the same parameters as for above-ground biomass. The fine roots will require roughly the same drying time as for leaves, whereas the large and medium-size roots will need a drying time equivalent to that of branches.

Regarding the stump, a subsample should be taken, preferably vertically to better take account of wood density variations in this part of the tree. Here, the same procedures should be used as for the disks of the trunk.

The calculations then made are the same as for above-ground biomass.